---
title: "Multi-Condition Analysis with CellDiff"
author: "Kai Guo"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Multi-Condition Analysis with CellDiff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette demonstrates how to use the CellDiff package for analyzing cell-cell communication differences across multiple conditions. CellDiff extends the functionality of CellChat by providing specialized tools for comparative analysis, with a focus on identifying and visualizing significant changes in cell-cell communication patterns.

We'll use the example dataset `cellchatlist` which contains three CellChat objects representing different genetic conditions: Wild Type (WT), Knockout (KO), and Double Knockout (DKO). Throughout this tutorial, we'll use WT as the reference condition.

## Setup

First, load the necessary packages:

```{r setup}
library(CellDiff)
library(ggplot2)
library(ComplexHeatmap)
```

Load the example dataset:

```{r load-data}
data(cellchatlist)

# Inspect the structure of the list
names(cellchatlist)
```

## 1. Multi-condition Pathway Comparison

One of the first steps in comparing multiple conditions is to identify signaling pathways that show significant differences. CellDiff provides the `rankDiffM` function for this purpose.

### Ranking Differential Pathways

```{r rank-pathways}
# Compare all conditions to WT as reference
pathway_results <- rankDiffM(
  object.list = cellchatlist,
  comparison_method = "all_vs_ref",  # Compare all conditions to the reference
  reference = "WT",                  # Set WT as reference
  use_log2fc = TRUE,                 # Use log2 fold change for better visualization
  comparison_plot_type = "barplot",  # Type of comparison plot (options: "barplot", "heatmap", "slope")
  show_comparison_barplot = TRUE,    # Show faceted barplot
  show_comparison_heatmap = TRUE,    # Show heatmap
  show_comparison_slope = TRUE,      # Show slope plot
  top.n = 10                         # Show top 10 differential pathways
)
```

### Visualizing Pathway Differences

The `rankDiffM` function returns various visualizations. Let's examine them:

```{r pathway-viz, fig.height=8, fig.width=10}
# Barplot showing pathway differences across conditions
pathway_results$comparison_barplot

# Heatmap showing pathway changes relative to WT
pathway_results$comparison_heatmap

# Slope plot showing pathway trends across conditions
pathway_results$comparison_slope_plot
```

### Extracting Top Differential Pathways

We can extract the top differential pathways for further analysis:

```{r top-pathways}
# Get the top differential pathways
top_pathways <- pathway_results$top_paths
head(top_pathways)

# Examine detailed statistics for the top pathways
pathway_stats <- pathway_results$data
head(pathway_stats[pathway_stats$name %in% top_pathways, c("name", "fold_change", "padj", "comparison")])
```

## 2. Heatmap Visualization of Differential Signaling

The `heatDiffM` function creates heatmaps showing differences in signaling patterns across cell types and pathways for multiple conditions.

### Basic Heatmap Comparison

```{r heatmap-basic, fig.height=8, fig.width=10}
# Create a multi-condition heatmap comparing sender signaling
heatmap_result <- heatDiffM(
  object.list = cellchatlist,
  comparison = c(1, 2, 3),       # Compare all three conditions
  reference = "WT",              # Set WT as reference
  measure = "sender",            # Focus on sender signaling
  use_log2fc = TRUE,             # Use log2 fold change
  cluster_rows = TRUE,           # Cluster cell types
  cluster_cols = TRUE,           # Cluster pathways
  show_values = TRUE,            # Show values in cells
  color.heatmap = c("blue", "white", "red")  # Custom colors
)

# Display the heatmap
heatmap_result$heatmap
```

### Customized Heatmap with Selected Pathways

```{r heatmap-custom, fig.height=8, fig.width=10}
# Create a custom heatmap with specific pathways
custom_heatmap <- heatDiffM(
  object.list = cellchatlist,
  comparison = c(1, 2, 3),
  reference = "WT",
  measure = "both",             # Both sender and receiver
  signaling = top_pathways[1:5], # Use top 5 differential pathways
  transpose = TRUE,             # Transpose the heatmap
  use_log2fc = TRUE,            # Use log2 fold change
  color.heatmap = c("purple", "white", "green"),  # Custom colors
  cluster_rows = TRUE,          # Cluster rows
  cluster_cols = TRUE,          # Cluster columns
  show_values = TRUE,           # Show values in cells
  value_digits = 1              # 1 decimal place
)

# Display the heatmap
custom_heatmap$heatmap
```

## 3. Multi-condition Scatter Plot Analysis

The `scatterDiffM` and `scatterDiff2DM` functions visualize how cell types change their signaling roles across conditions.

### 1D Scatter Plot Comparison

```{r scatter-1d, fig.height=7, fig.width=9}
# Compare signaling influence across conditions
scatter_result <- scatterDiffM(
  object.list = cellchatlist,
  comparison = c(1, 2, 3),       # Use all three conditions
  reference = "WT",              # Set WT as reference
  plot.type = "facet",           # Create a faceted plot
  comparison_method = "all_vs_ref",  # Compare all to reference
  measure.type = "influence",    # Use influence measure (sender * receiver)
  point.size = 4,                # Larger points
  point.alpha = 0.8,             # More opacity
  show_group_labels = TRUE,      # Show group labels
  label.size = 3.5               # Larger labels
)

# Display the plot
scatter_result
```

### 2D Scatter Plot Comparison

The 2D scatter plot shows changes in both sender and receiver roles:

```{r scatter-2d, fig.height=7, fig.width=9}
# For more detailed 2D visualization
scatter_result_2d <- scatterDiff2DM(
  object.list = cellchatlist,
  comparison = c(1, 2, 3),       # Use all three conditions
  reference = "WT",              # Set WT as reference
  plot.type = "facet",           # Create a faceted plot
  comparison_method = "all_vs_ref",  # Compare all to reference
  show.group.legend = TRUE,      # Show legend
  convex.hull = TRUE,            # Draw convex hulls
  highlight.cells = c("Cell1", "Cell2")  # Highlight specific cell types
)

# Display the plot
scatter_result_2d
```

## 4. Ligand-Receptor Pair Contribution Analysis

The `ContriDiffM` function examines the contributions of specific ligand-receptor pairs to signaling pathways across conditions.

```{r lr-contrib, fig.height=8, fig.width=10}
# Compare L-R pair contributions for the Pathway1 across conditions
lr_result <- ContriDiffM(
  object.list = cellchatlist,
  signaling = "Pathway1",
  reference = "WT",
  stack.method = "side-by-side",  # Display conditions side by side
  show.heatmap = TRUE,            # Show heatmap visualization
  top.n = 10                      # Limit to top 10 L-R pairs
)

# Display the barplot
lr_result$barplot

# Display the heatmap
if (!is.null(lr_result$heatmap)) {
  lr_result$heatmap
}
```

## 5. Finding Common and Unique Signaling Patterns

The `findCommonUniquePatterns` function identifies signaling patterns that are shared across conditions or unique to specific conditions.

```{r patterns}
# Identify common and unique patterns
patterns <- findCommonUniquePatterns(
  object.list = cellchatlist,
  comparison = c(1, 2, 3),       # Compare all three conditions
  min_overlap = 2,               # Minimum number of conditions to consider a pattern common
  return_networks = TRUE         # Return full network data
)

# Examine common patterns across conditions
head(patterns$common_patterns)

# Examine unique patterns for each condition
head(patterns$unique_patterns)
```

## 6. Advanced Network Visualization

CellDiff provides two main functions for network visualization: `netDiff` for chord diagrams and `networkLRDiff` for integrated network visualizations.

### Network Difference Visualization

```{r netdiff, fig.height=8, fig.width=8}
# Visualize network differences for a specific pathway
net_diff <- netDiff(
  object.list = cellchatlist,
  comparison = c(1, 2),         # Compare WT and KO
  signaling = "Pathway1",       # Focus on Pathway1 signaling
  use_normalized = TRUE,        # Normalize networks before comparison
  color.use = c("darkorange", "deepskyblue"),  # Custom colors
  link_style = "bezier",        # Use bezier curves
  highlight_sources = c("Cell1", "Cell2")  # Highlight specific sources
)
```

### Integrated Network Visualization

```{r network-lr, fig.height=8, fig.width=9}
# Integrated network visualization
network_viz <- networkLRDiff(
  object.list = cellchatlist,
  comparison = c(1, 2),         # Compare WT and KO
  pathways = top_pathways[1:3], # Use top 3 differential pathways
  node.size.factor = 1.2,       # Larger nodes
  edge.width.factor = 1.5,      # Thicker edges
  label.cell = TRUE,            # Label cell types
  node.label.repel = TRUE,      # Avoid label overlaps
  title = "WT vs KO Network Comparison"  # Custom title
)
```

## 7. Comprehensive Comparison of Multiple CellChat Objects

The `compareCellChatsM` function provides a global overview of changes in communication patterns across multiple conditions.

```{r global-comparison, fig.height=8, fig.width=9}
# Compare multiple CellChat objects to identify global changes
global_comparison <- compareCellChatsM(
  object.list = cellchatlist,
  comparison = c(1, 2, 3),      # Compare all three conditions
  reference = "WT",             # Set WT as reference
  measure.type = "sender",      # Focus on sender signaling
  show.heatmap = TRUE,          # Generate heatmap
  show.barplot = TRUE           # Generate barplot
)

# View the timeline showing changes across conditions
global_comparison$timeline
```

## Conclusion

This vignette has demonstrated how to use CellDiff for analyzing cell-cell communication differences across multiple conditions. The package provides a comprehensive set of tools for identifying and visualizing significant changes in intercellular signaling, helping researchers gain insights into the biological mechanisms underlying different experimental conditions.

Key functionalities demonstrated include:
- Identifying and ranking differential signaling pathways
- Visualizing signaling pattern changes with heatmaps
- Analyzing how cell types change their signaling roles
- Examining ligand-receptor pair contributions
- Finding common and unique signaling patterns
- Creating network visualizations

For more details about specific functions, please refer to the package documentation or visit the GitHub repository: https://github.com/guokai8/CellDiff.
